---
title: "Isolating Data with dplyr"
author: Adapted by Mauro Lepore from [RStudio primers](https://github.com/rstudio-education/primers) developed by [Garrett Grolemund](https://github.com/garrettgman)
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: default
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(fgeo.data)

checker <- function(label,
                    user_code,
                    check_code,
                    envir_result,
                    evaluate_result,
                    ...) {
  list(message = check_code,
  correct = TRUE,
  location = "append")
}

tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(echo = FALSE)
```

```{r stem-count}
stem_count <- fgeo.data::luquillo_stem_random %>% 
  filter(!is.na(dbh)) %>%
  add_count(CensusID, sp) %>% 
  select(CensusID, sp, n, everything())
```

## Welcome

In this case study, you will explore the abundance of tree-stems of different species over time. Along the way, you will master some of the most useful functions for isolating variables, cases, and values within a data frame:

* `select()` and `filter()`, which let you extract rows and columns from a data frame.
* `arrange()`, which lets you reorder the rows in your data.
* `%>%`, which organizes your code into reader-friendly "pipes".

This tutorial uses the [core tidyverse packages](http://tidyverse.org/), including __ggplot2__, __tibble__, and __dplyr__, as well as the __fgeo.data__ package. All of these packages have been pre-installed and pre-loaded for your convenience.

Click the Next Topic button to begin.

### The history of a species

Consider the dataset `stem_count`, containing a count of alive stems of different species recorded in censuses 1-6 of the Luquillo plot, in Puerto Rico. (`stem_count` is based on a small, random subset of trees and therefore it is OK for examples but invalid for research.)

```{r, echo=FALSE}
stem_count
```

You can use the data in `stem_count` to make graphs like this, which reveal the history of stems of a particular species.

```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%"}
stem_count %>% 
  filter(sp == "PREMON") %>%
  ggplot(aes(CensusID, n)) + 
  geom_line() +
  labs(title = "The history of PREMON")
```

But before you do, you will need to trim down `stem_count`. At the moment, there are more rows in `stem_count` than you need to build your plot. 

### An example

To see what I mean, consider how I made the plot above: I began with the entire data set, which if plotted as a scatterplot would've looked like this.

```{r out.width = "60%", cached = TRUE}
stem_count %>% 
  ggplot(aes(CensusID, n)) + 
  geom_point() +
  labs(title = "The history of every species in the dataset")
```

I then narrowed the data to just the rows that contain a single species, before plotting the data with a line geom. Here's how the rows with just one species look as a scatterplot.

```{r out.width = "60%", cached = TRUE}
stem_count %>% 
  filter(sp == "PREMON") %>%
  ggplot(aes(CensusID, n)) + 
  geom_line() +
  labs(title = "The history of PREMON")
```

If I had skipped this step, my line graph would've connected all of the points in the large data set, creating an uninformative graph.

```{r out.width = "60%", cached = TRUE}
stem_count %>% 
  ggplot(aes(CensusID, n)) + 
  geom_line() +
  labs(title = "The history of PREMON")
```

Your goal in this section is to repeat this process for PREMON (or a species that you choose). Along the way, you will learn a set of functions that isolate information within a data set. 

### Isolating data

This type of task occurs often in Data Science: you need to extract data from a table before you can use it. You can do this task quickly with three functions that come in the **dplyr** package:

1. **select()** - which extracts columns from a data frame.
1. **filter()** - which extracts rows from a data frame.
1. **arrange()** - which moves important rows to the top of a data frame.

Each function takes a dataframe or tibble as it's first argument and returns a new dataframe or tibble as its output. 

## select()

`select()` extracts columns of a data frame and returns the columns  as a new data frame. To use `select()`, pass it the name of a data frame to extract columns from, and then the names of the columns to extract. The column names do not need to appear in quotation marks or be prefixed with a `$`; `select()` knows to find them in the dataframe that you supply.

### Exercise - select()

Use the example below to get a feel for `select()`. Can you extract just the `CensusID` column? How about the `CensusID` and `sp` columns?

```{r select-1, exercise = TRUE, exercise.eval = TRUE, exercise.setup = "stem-count"}
select(stem_count, CensusID)
```

```{r select-1-solution}
select(stem_count, CensusID)
select(stem_count, CensusID, sp)
```

### select() helpers

You can also use a series of helpers with `select()`. For example, if you place a minus sign before a column name, `select()` will return every column but that column. Can you predict how the minus sign will work here?

```{r select-2, exercise = TRUE, exercise.setup = "stem-count"}
select(stem_count, -c(CensusID, sp))
```

The table below summarizes the other `select()` helpers that are available in __dplyr__. Study it, and then click "Continue" to test your understanding.

Helper Function   | Use                                               | Example
----------------- | ------------------------------------------------- | -------
**-**             | Columns except                               | `select(stem_count, -CensusID)`
**:**             | Columns between (inclusive)                  | `select(stem_count, CensusID:n)`
**contains()**    | Columns that contains a string               | `select(stem_count, contains("ID"))`
**ends_with()**   | Columns that ends with a string              | `select(stem_count, ends_with("ID"))`
**matches()**     | Columns that matches a regex                 | `select(stem_count, matches("ID"))`
**num_range()**   | Columns with a numerical suffix in the range | Not applicable with `stem_count`
**one_of()**      | Columns whose name appear in the given set     | `select(stem_count, one_of(c("CensusID", "treeID")))`
**starts_with()** | Columns that starts with a string            | `select(stem_count, starts_with("tree"))`

### select() quiz

```{r select-quiz}
question(
  "Which of these is not a way to select the `sp` and `n` columns together?",
  answer("select(stem_count, -c(sp, n))", , correct = TRUE),
  answer("select(stem_count, sp, n)"),
  answer("select(stem_count, CensusID:n)"),
  allow_retry = TRUE
)
```

## filter()

`filter()` extracts rows from a dataframe and returns them as a new dataframe. As with `select()`, the first argument of `filter()` should be a dataframe to extract rows from. The arguments that follow should be logical tests; `filter()` will return every row for which the tests return `TRUE`.

### filter in action

For example, the code chunk below returns every row with the `sp` value "PREMON" in `stem_count`. 

```{r echo = TRUE}
filter(stem_count, sp == "PREMON")
```

### Logical tests

To get the most from filter, you will need to know how to use R's logical test operators, which are summarised below.

Logical operator | tests                                | Example
---------------- | ------------------------------------ | ---------
**>**            | Is `x` greater than `y`?             | `x > y`
**>=**           | Is `x` greater than or equal to `y`? | `x >= y`
**<**            | Is `x` less than `y`?                | `x < y`
**<=**           | Is `x` less than or equal to `y`?    | `x <= y`
**==**           | Is `x` equal to `y`?                 | `x == y`
**!=**           | Is `x` not equal to `y`?             | `x != y`
**is.na()**      | Is `x` an `NA`?                      | `is.na(x)`
**!is.na()**     | Is `x` not an `NA`?                  | `!is.na(x)`

### Exercise - Logical Operators

See if you can use the logical operators to manipulate our code below to show:

* All stems in `stem_count` where n is greater than or equal to 230.
* All of the stems of the "CASARB" species.
* All of the stems that have a missing value for `n` (Hint: this should return an empty data set).

```{r filter-0, exercise = TRUE, exercise.setup = "stem-count"}
filter(stem_count, sp == "PREMON")
```

```{r filter-0-solution}
filter(stem_count, n >= 230)
filter(stem_count, sp == "CASARB")
filter(stem_count, is.na(n))
```

### Two common mistakes

When you use logical tests, be sure to look out for two common mistakes. One appears in each code chunk below. Can you find them? When you spot a mistake, fix it and then run the chunk to confirm that it works.

```{r eval = FALSE, echo = TRUE}
filter(stem_count, sp = "PREMON")
```

```{r filter-1a, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r filter-1a-solution}
filter(stem_count, sp == "PREMON")
```

```{r filter-1a-check}
"Good Job! Remember to use == instead of = when testing for equality."
```

```{r eval = FALSE, echo = TRUE}
filter(stem_count, sp == PREMON)
```

```{r filter-1b, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r filter-1b-solution}
filter(stem_count, sp == "PREMON")
```

```{r filter-1b-check}
"Good Job! As written this code would check that name is equal to the contents of the object named PREMON, which does not exist."
```

### Two mistakes - Recap

When you use logical tests, be sure to look out for these two common mistakes:

1. using `=` instead of `==` to test for equality.
2. forgetting to use quotation marks when comparing strings, e.g. `sp == PREMON`, instead of `sp == "PREMON"` 

### Combining tests

If you provide more than one test to `filter()`, `filter()` will combine the tests with an **and** statement (`&`): it will only return the rows that satisfy all of the tests. 

To combine multiple tests in a different way, use R's Boolean operators. For example, the code below will return all of the species encoded as PREMON or Sea or Anemone.

```{r echo = TRUE}
filter(stem_count, sp == "PREMON" | sp == "CASARB")
```

### Boolean operators

You can find a complete list or base R's boolean operators in the table below.

Boolean operator | represents                                 | Example
---------------- | ------------------------------------------ | ---------
**&**            | Are _both_ `A` and `B` true?               | `A & B`
`|`            | Are _one or both_ of `A` and `B` true?     | `A | B`
**!**            | Is `A` _not_ true?                         | `!A`
**xor()**        | Is _one and only one_ of `A` and `B` true? | `xor(A, B)`
**%in%**         | Is `x` in the set of `a`, `b`, and `c`?    | `x %in% c(a, b, c)`
**any()**        | Are _any_ of `A`, `B`, or `C` true?        | `any(A, B, C)`
**all()**        | Are _all_ of `A`, `B`, or `C` true?        | `all(A, B, C)`

### Exercise - Combining tests

Use Boolean operators to alter the code chunk below to return only the rows that contain:

* Stems from the fourth census of the PREMON species.
* Stems from the fourth or sixth censuses of the PREMON species.
* Stems which species are one of CORBOR, MICPRA, or EUGSTA.

```{r filter-2, exercise = TRUE, exercise.eval = TRUE, exercise.setup = "stem-count"}
filter(stem_count, sp == "PREMON" | sp == "CASARB")
```

```{r filter-2-solution}
filter(stem_count, CensusID == 4, sp == "PREMON")
filter(stem_count, CensusID == 4 | CensusID == 6, sp == "PREMON")
filter(stem_count, sp %in% c("CORBOR", "MICPRA", "EUGSTA"))
```

### Two more common mistakes

Logical tests also invite two common mistakes that you should look out for. Each is displayed in a code chunk below, one produces an error and the other is needlessly verbose. Diagnose the chunks and then fix the code.

```{r eval = FALSE, echo = TRUE}
filter(stem_count, 100 < n < 200)
```

```{r filter-3, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r filter-3-solution}
filter(stem_count, 100 < n, n < 200)
```

```{r filter-3-check}
"Good job! You cannot combine two logical tests in R without using a Boolean operator (or at least a comma between filter arguments)."
```

```{r eval = FALSE, echo = TRUE}
filter(stem_count, n == 50 | n == 60 | n == 70 | n == 80 | n == 90)
```

```{r filter-4, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r filter-4-solution}
filter(stem_count, n %in% c(50, 60, 70, 80, 90))
```

```{r filter-4-check}
"Good job! Although the first code works, you should make your code more concise by collapsing multiple or statements into an %in% statement when possible."
```

### Two more common mistakes - Recap

When you combine multiple logical tests, be sure to look out for these two common mistakes:

1. Collapsing multiple logical tests into a single test _without using a boolean operator_ 
1. Using repeated `|` instead of `%in%`, e.g. `x == 1 | x == 2 | x == 3` instead of `x %in% c(1, 2, 3)`


## arrange()

`arrange()` returns all of the rows of a dataframe reordered by the values of a column. As with `select()`, the first argument of `arrange()` should be a dataframe and the remaining arguments should be the names of columns. If you give `arrange()` a single column name, it will return the rows of the data frame reordered so that the row with the lowest value in that column appears first, the row with the second lowest value appears second, and so on. If the column contains character strings, `arrange()` will place them in alphabetical order.

### Exercise - arrange()

Use the code chunk below to arrange `stem_count` by `n`. Can you tell what the smallest value of `n` is? 

```{r arrange-1, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r arrange-1-solution}
arrange(stem_count, n)
```

```{r arrange-1-check}
"Good job!"
```

### Tie breakers

If you supply additional column names, `arrange()` will use them as tie breakers to order rows that have identical values in the earlier columns. Add to the code below, to make `sp` a tie breaker. The result should first order rows by value of `n` and then reorder rows within each value of `n` by values of `treeID`. 

```{r arrange-2, exercise = TRUE, exercise.eval = TRUE}
arrange(stem_count, n)
```

```{r arrange-2-solution}
arrange(stem_count, n, treeID)
```

### desc

If you would rather arrange rows in the opposite order, i.e. from _large_ values to _small_ values, surround a column name with `desc()`. `arrange()` will reorder the rows based on the largest values to the smallest.

Add a `desc()` to the code below to display the most abundant species for the latest census (the largest `CensusID` in the dataset) instead of the least abundant species.

```{r arrange-3, exercise = TRUE, exercise.eval = TRUE}
arrange(stem_count, n, desc(treeID))
```

```{r arrange-3-solution}
arrange(stem_count, desc(n), desc(treeID))
```

Think you have it? Click Continue to test yourself.

### arrange() quiz

Which species was the most abundant for a single quadrat in a single census? In the code chunk below, use `arrange()` to make the row with the smallest `treeID` value and the largest value of `n` appear at the top of the data set.

```{r arrange-4, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r arrange-4-solution}
arrange(stem_count, treeID, desc(n))
```

Now arrange `stem_count` so that the row with the largest value of `n` appears at the top of the dataframe. Will this be the same row? Why or why not?

```{r arrange-5, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r arrange-5-solution}
arrange(stem_count, desc(n))
```

## %>%

### Steps

Notice how each __dplyr__ function takes a dataframe as input and returns a dataframe as output. This makes the functions easy to use in a step by step fashion. For example, you could:

1. Filter `stem_count` to just pick stems which `stemID` is "32585".
2. Select the `CensusID`, `sp`, and `dbh` columns from the result.
3. Arrange those columns so that the largest `dbh` values appears near the top.

```{r echo = TRUE}
id_32585 <- filter(stem_count, stemID == "32585")
id_32585 <- select(id_32585, CensusID, sp, dbh)
id_32585 <- arrange(id_32585, desc(dbh))
id_32585
```

### Redundancy

The result shows us the largest stem was recorded in the sixth census, which was the latest in this dataset. But take a look at the code. Do you notice how we re-create `id_32585` at each step so we will have something to pass to the next step? This is an inefficient way to write R code.

You could avoid creating `id_32585` by nesting your functions inside of each other, but this creates code that is hard to read:

```{r echo = TRUE, eval = FALSE}
arrange(select(filter(stem_count, stemID == "32585"), CensusID, sp, dbh), desc(dbh))
```

The __dplyr__ package provides a third way to write sequences of functions: the pipe.

### %>%

The pipe operator `%>%` performs an extremely simple task: it passes the result on its left into the first argument of the function on its right. Or put another way, `x %>% f(y)` is the same as `f(x, y)`. This piece of code punctuation makes it easy to write and read series of functions that are applied in a step by step way. For example, we can use the pipe to rewrite our code above:

```{r echo = TRUE}
stem_count %>% 
  filter(stemID == "32585") %>% 
  select(CensusID, sp, dbh) %>% 
  arrange(desc(dbh))
```

As you read the code, pronounce `%>%` as "then". You'll notice that __dplyr__ makes it easy to read pipes. Each function name is a verb, so our code resembles the statement, "Take `stem_count, _then_ filter it by `stemID`, _then_ select the `CensusID` and `dbh` columns, _then_ arrange the results by descending values of `dbh`."

__dplyr__ also makes it easy to write pipes. Each __dplyr__ function returns a dataframe that can be piped into another __dplyr__ function, which will accept the dataframe as its first argument. In fact, __dplyr__ functions are written with pipes in mind: each function does one simple task. __dplyr__ expects you to use pipes to combine these simple tasks to produce sophisticated results.

### Exercise - Pipes

I'll use pipes for the remainder of the tutorial, and I will expect you to as well. Let's practice a little by writing a new pipe in the chunk below. The pipe should:

1. Filter `stem_count` to just pick stems which `stemID` is _"138222"_.
2. Select the `CensusID`, `sp`, and `dbh` columns from the result.
3. Arrange those columns so that the largest `dbh` values appears near the top.

Try to write your pipe without copying and pasting the code from above.

```{r pipe-1, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r pipe-1-solution}
stem_count %>% 
  filter(stemID == "138222") %>% 
  select(CensusID, sp, dbh) %>% 
  arrange(desc(dbh))
```

### Your name

You've now mastered a set of skills that will let you easily plot the different species over time. In the code chunk below, use a combination of __dplyr__ and __ggplot2__ functions with `%>%` to:

1. Trim `stem_count` to just the rows that contain the PREMON species and your 
2. Trim the result to just the columns that will appear in your graph (not strictly necessary, but useful practice)
3. Plot the results as a line graph with `year` on the x axis and `prop` on the y axis

Note that the first argument of `ggplot()` takes a data frame, which means you can add `ggplot()` directly to the end of a pipe. However, you will need to switch from `%>%` to `+` to finish adding layers to your plot.

```{r your-name-1, exercise = TRUE, exercise.setup = "stem-count"}

```

```{r your-name-1-solution}
stem_count %>% 
  filter(sp == "PREMON") %>%
  ggplot(aes(CensusID, n)) + 
  geom_line() +
  labs(title = "The history of PREMON")
```

### Recap

Together, `select()`, `filter()`, and `arrange()` let you quickly find information displayed within your data.

The next tutorial will show you how to derive information that is implied by your data, but not displayed within your data set.

In that tutorial, you will continue to use the `%>%` operator, which is an essential part of programming with the 

Pipes help make R expressive, like a spoken language. Spoken languages consist of simple words that you combine into sentences to create sophisticated thoughts.

In the tidyverse, functions are like words: each does one simple task well. You can combine these tasks into pipes with `%>%` to perform complex, customized procedures.
